---
title: "Week 4:Simulation & Profiling"
author: "Katherine Ramírez Cubillos"
date: "23/8/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# The srt Function

**srt: Compactly display the internal structure of an R object**

- A diagnostic funciton and an alternative to 'summary'. 
- It is especially well studied to compactly display the (abbreviated) contents of (possibily nested) lists. 
- Roughly one line per basic object.
- The basic goal of str is to answer the question: What is in the object?
```{r}
str(str)
str(lm)
```
It shows the funciton arguments for the **lm** function. You can see a very brief summary, it takes the first argument's a formula, the second argument's data, etc. 

The next example generates some normal random variables, 100 of the, mean two variant, and standard derivation four.  
```{r}
x <- rnorm(100, 2, 4)
summary(x)
## give the first five numbers in the vector x
str(x)

# The factor has 40 levels and each one is 
# repeated ten times so if I call str on it. 
f <- gl(40,10)
str(f)
# The output's a llittle bit different. 
# It gives the number of elements in each of the 40 different levels. 
summary(f)

```
It is possible to call str to get a little some different output. In a data-frame it tells that there's a 153 observations, so there are 153 rows in the data frame, with six variables and then for each variable it gives a little output. 

```{r}
library(datasets)
head(airquality)
str(airquality) 
#Quick examination of data that you might have in R. 
# And what the structure of different R objects is. 
```
Matrix with random normals in there. It will be a 10 by 10 matrix. The function srt will give a little bit more information. It'll say that it's a two-dimensional array. 
```{r}
m <- matrix(rnorm(100), 10,10)
str(m)
```
The next example is a list by using the split funciton and see how srt can look at the list and give a compact summary of it. It takes the airquality data frame and split it by the month. 
```{r}
s <- split(airquality, airquality$Month)
str(s)
```
This list contains five different data frames where each data frame corresponds to the data for a given month. 

# Simulation

Funcitons for probability distribution in R:

- **rnorm**: generate random Normal variates with a given mean and standar deviation. 
- **dnorm**: evaluate the Normal probability density (with a given mean/SD) at a point (or vector of points)
- **pnorm**: evaluate the cumulative distribution function for a normal distribution. 
- **rpois** generate random Poisson variates with a given rate. 

## Generating Random Numbers

Probability distribution functions usually have four funcitons associated with them. The function are prefixed with a:

- **d** for density.
- **r** for random number generation. 
- **p** for cumulative distribution. 
- **q** for quantile funciton. 

Working with the Normal distributions requires using four functions. 
```{r}
dnorm(x, mean = 0, sd = 1, log = FALSE)
# pnorm(q, mean = 0, sd = 1, lower.tail = TRUE, log.p = FALSE)
# qnorm(p, mean = 0, sd = 1, lower.tail = TRUE, log.p = FALSE)
# rnorm(n, mean = 0, sd = 1)

```
If $\phi$ is the cumulative distribution for a standar Normal distribution, then $pnorm(q)=\Phi(q)$ and $qnorm(p)=\Phi^{-1}(p)$ 
```{r}
x <- rnorm(10)
x

x <- rnorm(10,20,2)
x
```
Setting the random number with **set.seed** ensures reproducibility. 

```{r}
set.seed(1)
rnorm(5)
rnorm(5)
set.seed(1)
rnorm(5)
```
Always set the random number seed when conducting a simulation!

## Generating Poisson data

```{r}
rpois(10,1)
rpois(10,2)
rpois(10,20)
ppois(2,2)   ## Cumulative distribution
             ## Pr(x <= 2)
ppois(4,2)   
             ## Pr(x <= 4)
ppois(6,2)  
             ## Pr(x <= 6)
summary(x)
```
## Generating Random Numbers From a Linear Model

Suppose we want to simulate from the following linear model:

$y=\beta_0 + \beta_1x + \epsilon$

where $\epsilon ~ N(0,2^2)$. Assume $x ~ N(0,1^2)$, $\beta_0 = 0.5$ and $\beta_1 = 2$ 

```{r}
set.seed(20)
x <- rnorm(100)
e <- rnorm(100, 0, 2) # std = 0 & media =2. 
y <- 0.5 + 2*x + e
summary(y)
plot(x,y)

```
What if **x** is binary?
```{r}
set.seed(10)
x <- rbinom(100, 1, 0.5)
e <- rnorm(100, 0, 2)
y <- 0.5 + 2*x + e
summary(y)
plot(x, y)
```
## Generating Random Numbers From a Generalized Linear Model

Suppose we want to simulate from a Poisson model where:

$Y=Poisson \ (u)$

$\log u = \beta_0 + \beta_1x$

And $\beta_0 = 0.5$ and $\beta_1 = 0.5$. We need to use the **rpois** function for this. 

```{r}
set.seed(1)
x <- rnorm(100)
log.mu <- 0.5 + 0.3*x
y <- rpois(100, exp(log.mu))
summary(y)
plot(x, y)
```
## Random Sampling

The **sample** function draws randomly from a specified set of (scalar) objects allowing you to sample from arbitrary distributions. 

```{r}
set.seed(1)
# I pass the vector of integers of one to ten. 
# I want o sample randomly four of them, without replacements. 
sample(1:10, 4)
sample(1:10, 4)
sample(letters, 5)
sample(1:10)   ## permutation
sample(1:10)
sample(1:10, replace=TRUE)  ## Sample w/replacement
```
### Simulation

- Drawing samples from specific probability distributions can be done with **r** functions. 
- Standar distributions are built in: Normal, Poisson, Binomial, Exponential, Gamma, etc. 
- The **sample** function can be used to draw random samples from arbitrary vectors. 
- Setting the random number generator seed via set.seed is critical for reproducibility.

## Profiling R Code

**Why is my code so slow?**

- Proffing is a systematic way to examine how much time is spend in different parts of a program. 
- Useful when trying to optimize your code. 
- Often code runs fine once, but what if you have to put it in a loop for 1,000 interation? Is it still fast enough?
- Profiling is better than guessing. 

**On optimizing your code**

- Getting biggest impact on speeding up code depends on knowing where the code spends most of its time. 
- This cannot be done without performance analysis or profiling. 
- We should forget about small efficiencies, say about 97% of the time: premature optimization is the root of all evil. (Donald Knuth)

### General principles of optimization

- Design first, then optimize. 
- Remember: Premature optimization is the root of evil. 
- Measure (collect data), don't guess. 
- If you're going to be scientist, you need to apply the same principles here!

### Using system.time()

- Usually, the user time and elapsed time are relatively close, for straight computing tasks. 
- Elapsed time may be *greater than* user time if the CPU spends a lot of time waiting around. 
- Elapsted time may be *smaller than* the user time if your machine has multiple cores/processors (and is capable of using them)
 - Multi-threaded BLAS libraries (vecLib/Accelerate, ATLAS, ACML, MKL)
 - Parallel processing via the **parallel** package. 
```{r}
## Elapsed time > user time
system.time(readLines("http://www.jhsph.edu"))

## Elapsed time < user time
hilbert <- function(n){
  i <- 1:n
  1 / outer(i - 1, i, "+")
}
x <- hilbert(1000)
system.time(svd(x))
```
### Timing longer expressions
```{r}
system.time({
  n <- 1000
  r <- numeric(n)
  for(i in 1:n){
    x <- rnorm(n)
    r[i] <- mean(x)
  }
})

```
### Beyond system.time()

- Using **system.time()** allows you to test certain functions or code blocks to see if they are taking excessive amounts of time. 
- Assumes you already know where the problem is and can call **system.time()** on it. 
- What if you don't know where to start?

### The R Profiler

- The **Rprof()** funciton starts the profiler in R. 
 - R must be compiled with profiler support (but this is usually the case)

-  The **summaryRprof()** funciotn summarizes the output from **Rprof()** (otherwise it's not readable)

 - Do not use **system.time()** and **Rprof()** together or you will be sad. 
 
 - **Rprof()** keeps track of the function call stack at regularly sample intervals and tabulates how much time is spend in each funciton. 

- Defult sampling interval is 0.02 seconds. 

- NOTE: If your code runs very quicly, the profiler is not useful, but then you probably don't need it in that case. 

### Using summaryRprof()

- The **summaryRprof()** funciton tabulates the R profiler output and calculates how much time is spend in which funciton. 
- There are two methods for normalizing the data. 
- "by.total" divides the time spend in each funciton by the total run time. 
- "by.self" does the same but first subtracts out time spent in functions above in the call stack. 

```{r}
## lm(y~x)
sample.interval = 10000
# $by.total
# $by.self
# $sample.interval
# $sample.time
```
### Summary

- **Rprof()** runs the profiler for performance of analysis of R code. 
- **summaryRprof()** summarizes the output of **Rprof()** and gives percent of time spent in each function (with two of normalization)
- Good to break your code into functions so that the profiler can give useful informacion about where time is being spent. 
- C or Fortran code is not profiled. 

```{r}
```
 